Tropical Mathematics for the Lambda Calculus

### 2023-01-18 section1.tex
commit

Tropical mathematics was introduced in the seventies by the Brazilian mathematician Imre Simon \cite{Simon} (whose origin stands behind the adjective ``tropical'') as a way to do algebra and geometry by replacing the usual ring structure of numbers based on addition and multiplication by the semi-ring structure given by ``$\min$'' and ``$+$''.
%
%
% interpreting the usual ``$\times$'' and  ``$+$'' operations by  ``$+$'' by ``$\min$''. It can thus be seen as a sort of ``logarithmic'' version of usual geometry (this idea can be made precise via the so-called \emph{Maslov deformation} \cite{}).
%Tropical mathematics is a form of \emph{idempotent} mathematics, since the role of addition is 
%played by the idempotent operation $\min$.
For instance, a quadratic polynomial like e.g.~$p(x)=ax^{2}+bx+c$, when interpreted over the tropical semi-ring, becomes the piecewise linear function
$
\varphi(\alpha)=\min\{2\alpha + a, \alpha+b, c\}
$.

As working on the idempotent $\min,+$-semiring is often easier that working with addition and multiplication, the tropicalization of  algebraic concepts makes several intractable properties (e.g.~finding roots of polynomials, or studying geometric invariants like e.g.~the genus) become tractable. For this reason, methods from tropical geometry are more and more applied in computer science, e.g.~for convex analysis and machine learning (see \cite{Maragos2021} for a recent survey).

>>>>

Tropical mathematics was introduced in the seventies by the Brazilian mathematician Imre Simon \cite{Simon}
 (whose origin stands behind the adjective ``tropical'') 
 as a combinatorial approach to algebra and geometry obtained by replacing the usual ring structure of numbers based on addition and multiplication by the semi-ring structure given by ``$\min$'' and ``$+$''.
%
%
% interpreting the usual ``$\times$'' and  ``$+$'' operations by  ``$+$'' by ``$\min$''. It can thus be seen as a sort of ``logarithmic'' version of usual geometry (this idea can be made precise via the so-called \emph{Maslov deformation} \cite{}).
%Tropical mathematics is a form of \emph{idempotent} mathematics, since the role of addition is 
%played by the idempotent operation $\min$.
For instance, a quadratic polynomial like e.g.~$p(x)=ax^{2}+bx+c$, when interpreted over the tropical semi-ring, becomes the piecewise linear function
$
\varphi(\alpha)=\min\{2\alpha + a, \alpha+b, c\}
$.

Working with $\min$ and $+$ is generally easier than working with standard addition and multiplication; for instance, one of the fundamental and most intractable problems in algebra, that of finding roots of polynomials, admits a linear time algorithm in the tropical case (and the roots of a tropicalized polynomial can be used to approximate the actual roots of the polynomial \cite{Noferini2015}). Such a computational nature of tropical concepts explains why these are widely applied in computer science, e.g.~for convex analysis and machine learning (see \cite{Maragos2021} for a recent survey).

### 2023-01-18 section5.tex
commit

Consider now the following term $M$ of type $\bool$:
\[
 (\true \oplus_p\false)\oplus_p((\true\oplus_p\false)\oplus_p(\true\oplus_p\false))
\]
Let us give addresses $\omega\in\set{00,01,100,101,110,111}$ to the occurrences of $\true,\false$ in $M$, by following the tree structure of $M$, reading $0$ as ``left'' and $1$ as ``right''.
The same addresses also represent all the different possible reduction paths from $M$ to a normal form.
For instance, $100$ represents the reduction which keeps the right part of the outermost $\oplus_p$ and erases the left part, then continues by choosing the left part twice, reaching at the end the occurrence $\true_{100}$ in $M$, i.e.\ the second occurrence of $\true$ in $M$, starting from the left.
Calling $q:=1-p$, there are the following six normal terms reachable from $M$:
$P_{00}(p,q)\true$, 
$P_{100}(p,q)\true$, 
$P_{111}(p,q)\true$, 
$P_{01}(p,q)\false$, 
$P_{110}(p,q)\false$,
$P_{101}(p,q)\false$,
where the $P$'s are the following monomials in $p,q$:
$P_{00}(p,q):=p^2$,
$P_{100}(p,q):=qp^2$,
$P_{111}(p,q):=q^3$,
$P_{01}(p,q):=pq$,
$P_{110}(p,q)=P_{101}(p,q):=q^2p$.
They correspond to the respective reduction path from $M$ to the normal term of the same address.
$P_{\omega}$ is then the probability (as a function of $p,q$) of obtaining the respective occurrence $\true_{\omega}$ or $\false_\omega$ in $M$ after reducing it to normal form.
Thinking, as we will, to $p,q$ as parameters, it can thus also be read as the \emph{likelihood function} of such event.
The polynomial $Q_{1}(p,q):=P_{00}(p,q)+P_{100}(p,q)+P_{111}(p,q)$ gives instead the whole probability of obtaining $\true$ after such samplings, and $Q_{0}(p,q):=P_{01}(p,q)+P_{110}(p,q)+P_{101}(p,q)$ the one for $\false$.
%Let us consider in this subsection a probabilistic extension of $\lam$-calculus, call it $\STLC_\oplus$, adding as usual terms of shape $pM+qN$ and $M\oplus_p N$, for $p,q\in[0,1]$.These terms are typed via the rule:
%\[
%\dfrac{\Gamma\vdash M:A \qquad \Gamma\vdash N:A}{\Gamma\vdash M\oplus_p N:A}
%\]
%and similar for $\Gamma\vdash pM+qN:A$.We add the reduction rule:
%\[
% M\oplus_p N \to pM+(1-p)N
%\]
%so that such terms play the role of probabilistic choices of parameter $p$, as well as the rule:
%\[
% pM+qM\to (p+q)M.
%\]
%Let us consider $M:=(I\oplus_p\Omega)\oplus_p((I\oplus_p\Omega)\oplus_p(I\oplus_p\Omega))$. Reducing to normal form, we have:
%\[
% M\twoheadrightarrow (p^2+(1-p)p^2+(1-p)^3)I+(p(1-p)+2(1-p)^2p)\Omega.
%\]
%The index $\omega\in\set{00,100,111,01,110,101}$ of each $P_\omega$ indicates the path of the reduction that led from $M$ to the respective occurrence $I_\omega$ of $I$ or $\Omega_\omega$ of $\Omega$ from $M$ to its normal form ($0$ means ``left'' and $1$ means ``right'').For instance, in order to reach $I_{100}$, i.e.\ the second occurrence of $I$ from the left in $M$, we have to take the right path in the outer $\oplus_p$ of $M$, then two times the left path in the new outer $\oplus_p$'s that we encounter during reduction.$P_{\omega}(p,q)$ gives then the probability (as a function of $p,q$) of obtaining the respective occurrence $I_{\omega}$ or $\Omega_\omega$ in the normal form, if we were to sample at each time we reduce a $\oplus_p$.It can thus also be read as the likelihood function of such event.The polynomials $Q_{1,2}(p,q)$ give instead the whole probability of obtaining respectively $I$ or $\Omega$, in the normal form after such samplings.

The tropical point of view allows now to express two natural questions about this situation:
\begin{enumerate}
 \item What is the \emph{maximum likelihood estimator} for the event ``$M$ produces $\true_\omega$'' (similarly for $\false_\omega$)?
 This means asking which is the choice of the parameters $p,q$ that maximizes the probability of getting $\true_\omega$ by repeatedly sampling $M$.
 \item Knowing that $M$ produced $\true$ (similarly for $\false$), which is the choice of the parameters $p,q$ that maximizes the probability that the produced occurrence of $\true$ was a fixed $\true_{\omega_0}$ (i.e. that among the paths leading to $\true$, the one taken was $\omega_0$)?
\end{enumerate}

These are questions much in the style of those asked in {\color{red}Ref?} in the framework of {\color{red}??}.

For 1), we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and $(p,q)$ belong to:
\begin{equation}
  \begin{array}{r@{}c@{}l@{}}
   \underset{(x,y)}{\RM{arg max}} P_\omega (x,y)
   & = &
   \underset{(x,y)}{\RM{arg min}}\set{-\log_c P_\omega (x,y)}
   \\
   & = &
   \underset{(x,y)}{\RM{arg max}}\set{(\trop P_\omega) (-\log_c x,-\log_c y)} \label{eq:argmax}
  \end{array}
\end{equation}
where the equalities hold for all $0<c\neq1$.

For 2), %the $\omega_0$ is s.t.\ $\max\limits_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$. So
we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and
$\max\limits_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$.
This last condition is equivalent to ask that:
\begin{equation}
  \begin{array}{l@{}l@{}}
    \min\limits_{\omega\in\set{00,100,111}} \, -\log_c P_\omega(x,y) = -\log_c P_{\omega_0}(x,y)
   & \textit{ i.e.: }
   \\
   (\trop Q_1)( -\log_c x, -\log_c y) = (\trop P_{\omega_0})( -\log_c x, -\log_c y). &  \label{eq:max}
  \end{array}
\end{equation}
In both cases, the tropicalisation operator $\trop{}$ as well as the \emph{negative $\log$-probabilities} appear.

\begin{remark}
 By adapting [Section IV, Section VI, \cite{Manzo2013}], it can be seen that $\LREL_!$ is a model of $\STLC_\oplus$.
 In particular, if we set $\model \bool:=\set{0,1}$, our running example $M$ is interpreted as $\model M\in\Lawv^{\set{0,1}}$ giving the tropicalised probabilities: $\model M_0=(\trop Q_0)(p,1-p)$, $\model M_1=(\trop Q_1)(p,1-p)$.
\end{remark}

%Now, in $\LREL$, seen as a model of such probabilistic $\lam$-calculus, the interpretation of a term already computes the tropicalisation of the polynomials expressing the probabilities, because the underlying semiring of the model is tropical.
%For instance, for our running example $M$:
%\[\model M = \min\left\{\trop{Q_1}(p,1-p) \cdot \model I, \trop{Q_0}(p,1-p) \cdot \model \Omega\right\}.\]
Therefore, assuming the interpretation of a term known and assuming the tropical polynomials (in $p$ and $q$, not after the substitution $q:=1-p$) are extracted from it, one could study the optimisation problems \ref{eq:argmax}, \ref{eq:max} \emph{without} having to reduce the term.

@@

\begin{remark}
 Taking the usual $M$ as example, the $p\in[0,1]$ s.t.\ $(p,1-p)$ is a tropical root of $\trop Q_1$ or of $\trop Q_0$ give, by definition of tropical root, the values of the probabilistic parameter in $\oplus_p$ for which there are at least two different equiprobable paths of samples from $M$ to its normal form.
 Moreover, looking at Equation~\ref{eq:max}, we see that the $p\in[0,1]$ s.t.\ $(-\log_c p,-\log_c(1-p))$ is a tropical root of, say, $\trop Q_1$ are the values of the probabilistic parameter in $\oplus_p$ for which there are at least two different equiprobable occurrences of $\true$ that are sampled by $M$ during its reduction to normal form, knowing that some $\true$ was sampled.
\end{remark}

>>>>

Consider now the following term $M$ of type $\bool$:
\[
 (\true \oplus_p\false)\oplus_p((\true\oplus_p\false)\oplus_p(\true\oplus_p\false))
\]
Let us give addresses $\omega\in\set{00,01,100,101,110,111}$ to the occurrences of $\true,\false$ in $M$, by following the tree structure of $M$, reading $0$ as ``left'' and $1$ as ``right''.
The same addresses also represent all the different possible reduction paths from $M$ to a normal form.
For instance, $100$ represents the reduction which keeps the right part of the outermost $\oplus_p$ and erases the left part, then continues by choosing the left part twice, reaching at the end the occurrence $\true_{100}$ in $M$, i.e.\ the second occurrence of $\true$ in $M$, starting from the left.
Calling $q:=1-p$, there are the following six normal terms reachable from $M$:
$P_{00}(p,q)\true$, 
$P_{100}(p,q)\true$, 
$P_{111}(p,q)\true$, 
$P_{01}(p,q)\false$, 
$P_{110}(p,q)\false$,
$P_{101}(p,q)\false$,
where the $P$'s are the following monomials in $p,q$:
$P_{00}(p,q):=p^2$,
$P_{100}(p,q):=qp^2$,
$P_{111}(p,q):=q^3$,
$P_{01}(p,q):=pq$,
$P_{110}(p,q)=P_{101}(p,q):=q^2p$.
They correspond to the respective reduction path from $M$ to the normal term of the same address.
$P_{\omega}$ is then the probability (as a function of $p,q$) of obtaining the respective occurrence $\true_{\omega}$ or $\false_\omega$ in $M$ after reducing it to normal form.
Thinking of $p,q$ as parameters, $P_{\omega}$ can thus be read as the \emph{likelihood function} of the event $\omega$.
The polynomial $Q_{1}(p,q):=P_{00}(p,q)+P_{100}(p,q)+P_{111}(p,q)$ gives instead the whole probability of obtaining $\true$ after such samplings, and $Q_{0}(p,q):=P_{01}(p,q)+P_{110}(p,q)+P_{101}(p,q)$ the one for $\false$.

%Let us consider in this subsection a probabilistic extension of $\lam$-calculus, call it $\STLC_\oplus$, adding as usual terms of shape $pM+qN$ and $M\oplus_p N$, for $p,q\in[0,1]$.These terms are typed via the rule:
%\[
%\dfrac{\Gamma\vdash M:A \qquad \Gamma\vdash N:A}{\Gamma\vdash M\oplus_p N:A}
%\]
%and similar for $\Gamma\vdash pM+qN:A$.We add the reduction rule:
%\[
% M\oplus_p N \to pM+(1-p)N
%\]
%so that such terms play the role of probabilistic choices of parameter $p$, as well as the rule:
%\[
% pM+qM\to (p+q)M.
%\]
%Let us consider $M:=(I\oplus_p\Omega)\oplus_p((I\oplus_p\Omega)\oplus_p(I\oplus_p\Omega))$. Reducing to normal form, we have:
%\[
% M\twoheadrightarrow (p^2+(1-p)p^2+(1-p)^3)I+(p(1-p)+2(1-p)^2p)\Omega.
%\]
%The index $\omega\in\set{00,100,111,01,110,101}$ of each $P_\omega$ indicates the path of the reduction that led from $M$ to the respective occurrence $I_\omega$ of $I$ or $\Omega_\omega$ of $\Omega$ from $M$ to its normal form ($0$ means ``left'' and $1$ means ``right'').For instance, in order to reach $I_{100}$, i.e.\ the second occurrence of $I$ from the left in $M$, we have to take the right path in the outer $\oplus_p$ of $M$, then two times the left path in the new outer $\oplus_p$'s that we encounter during reduction.$P_{\omega}(p,q)$ gives then the probability (as a function of $p,q$) of obtaining the respective occurrence $I_{\omega}$ or $\Omega_\omega$ in the normal form, if we were to sample at each time we reduce a $\oplus_p$.It can thus also be read as the likelihood function of such event.The polynomials $Q_{1,2}(p,q)$ give instead the whole probability of obtaining respectively $I$ or $\Omega$, in the normal form after such samplings.

This way the probabilistic evaluation of $M$ is presented in the form of a \emph{hidden Markov model} \cite{Baum1966}, a fundamental statistical model, and notably one to which tropical methods are generally applied \cite{Pachter2004}. Indeed, typical questions in this case would be:
%
%The tropical point of view allows now to express two natural questions about this situation:
\begin{enumerate}
 \item What is the \emph{maximum likelihood estimator} for the event ``$M$ produces $\true_\omega$'' (similarly for $\false_\omega$)?
 This means asking which is the choice of the parameters $p,q$ that maximizes the probability of getting $\true_\omega$ by repeatedly sampling $M$.
 \item Knowing that $M$ produced $\true$ (similarly for $\false$), which is the choice of the parameters $p,q$ that maximizes the probability that the produced occurrence of $\true$ was a fixed $\true_{\omega_0}$ (i.e. that among the paths leading to $\true$, the one taken was $\omega_0$)?
\end{enumerate}
A similar argument could be done by replacing $\true$ and $\false$ by, respectively, a converging and a diverging term (e.g.~in a $\B{PCF}$-style language), so 1) would be about finding maximum likelihood estimators for the event ``$M$ converges''.

%These are questions much in the style of those asked in {\color{red}Ref?} in the framework of {\color{red}??}.
Answering 1) and 2) amounts at solving a maximization problem related to $P_{\omega}$, which is more easily solved by 
passing to the tropical polynomial $\trop P_{\omega}$. 
For 1), we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and $(p,q)$ belong to:
\begin{equation}
  \begin{array}{r@{}c@{}l@{}}
   \underset{(x,y)}{\RM{arg max}} P_\omega (x,y)
   & = &
   \underset{(x,y)}{\RM{arg min}}\set{-\log_c P_\omega (x,y)}
   \\
   & = &
   \underset{(x,y)}{\RM{arg max}}\set{(\trop P_\omega) (-\log_c x,-\log_c y)} \label{eq:argmax}
  \end{array}
\end{equation}
where the equalities hold for all $0<c\neq1$.

For 2), %the $\omega_0$ is s.t.\ $\max\limits_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$. So
we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and
$\max\limits_{\omega\in\set{00,100,111}} \, P_\omega(x,y) = P_{\omega_0}(x,y)$.
This last condition is equivalent to ask that:
\begin{equation}
  \begin{array}{l@{}l@{}}
    \min\limits_{\omega\in\set{00,100,111}} \, -\log_c P_\omega(x,y) = -\log_c P_{\omega_0}(x,y)
   & \textit{ i.e.: }
   \\
   (\trop Q_1)( -\log_c x, -\log_c y) = (\trop P_{\omega_0})( -\log_c x, -\log_c y). &  \label{eq:max}
  \end{array}
\end{equation}
In both cases, passing to $\trop P_{\omega}$ makes the problem easier, as this may be reduced to computing the tropical roots of the polynomial (which can be done in linear time \cite{Noferini2015}). %the tropicalisation operator $\trop{}$ as well as the \emph{negative $\log$-probabilities} appear.

\begin{remark}
 By adapting [Section IV, \cite{Manzo2013}], it can be seen that $\LREL_!$ is a model of $\STLC_\oplus$.
 In particular, if we set $\model \bool:=\set{0,1}$, our running example $M$ is interpreted as $\model M\in\Lawv^{\set{0,1}}$ giving the tropicalised probabilities: $\model M_0=(\trop Q_0)(p,1-p)$, $\model M_1=(\trop Q_1)(p,1-p)$.
\end{remark}

%Now, in $\LREL$, seen as a model of such probabilistic $\lam$-calculus, the interpretation of a term already computes the tropicalisation of the polynomials expressing the probabilities, because the underlying semiring of the model is tropical.
%For instance, for our running example $M$:
%\[\model M = \min\left\{\trop{Q_1}(p,1-p) \cdot \model I, \trop{Q_0}(p,1-p) \cdot \model \Omega\right\}.\]
Therefore, assuming the interpretation of a term is known, 
so that one can extract the associated tropical polynomials from it,
% (in $p$ and $q$, not after the substitution $q:=1-p$) 
% are extracted from it, 
one could study the optimisation problems \ref{eq:argmax}, \ref{eq:max} \emph{without} having to actually reduce the term.

@@

\begin{remark}\label{rem:troproots}
 Taking the usual $M$ as example, the $p\in[0,1]$ s.t.\ $(p,1-p)$ is a tropical root of $\trop Q_1$ or of $\trop Q_0$ provide, by definition, the values of the probabilistic parameter in $\oplus_p$ for which there are at least two different equiprobable paths of samples from $M$ to its normal form.
 Moreover, looking at Equation~\ref{eq:max}, we see that the $p\in[0,1]$ s.t.\ $(-\log_c p,-\log_c(1-p))$ is a tropical root of, say, $\trop Q_1$ are the values of the probabilistic parameter in $\oplus_p$ for which there are at least two different equiprobable occurrences of $\true$ that are sampled by $M$ during its reduction to normal form, knowing that some $\true$ was sampled.
\end{remark}

### 2023-01-18 section8.tex
Mini mini commit
Ho levato la frase "In our opinion" dalla prima riga della conclusione. (Ma se la preferisci, rimettila pure)
commit

In our opinion, the main goals of this paper are two. Firstly,  to
demonstrate the existence of a conceptual bridge between two different well-studied quantitative approaches to higher-order programs, and to highlight the possibility of transferring results and techniques from one approach to the other. 
Secondly, to suggest that tropical mathematics, a
field which has been largely and successfully applied in computer science, could be used to study quantitative properties of higher-order programs.

While the first goal has been developed in detail, and at different levels of abstraction, for the second goal we only sketched a few interesting directions (best-case analysis, log-probabilities, differential privacy). We believe that exploring in more depth these ideas could be a fruitful direction; moreover, 
since both generalized metrics and quantale-modules have been largely applied in computer science, 
a natural question is whether the generalized approach of Section \ref{section6} could  lead to new applications of metric and tropical methods to the $\lambda$-calculus.

>>>>

%In our opinion,
The main goals of this paper are two. Firstly,  to
demonstrate the existence of a conceptual bridge between two different well-studied quantitative approaches to higher-order programs, and to highlight the possibility of transferring results and techniques from one approach to the other. 
Secondly, to suggest that tropical mathematics, a
field which has been largely and successfully applied in computer science, could be used to study quantitative properties of higher-order programs.

While the first goal was here developed in detail, and at different levels of abstraction, for the second goal we only sketched a few interesting directions (best-case analysis, log-probabilities, differential privacy). We believe that exploring these ideas in more depth could be a fruitful direction; moreover, 
since both generalized metrics and quantale-modules have been largely studied in computer science, 
a natural question is if the generalized approach of Section \ref{section6} could  lead to new applications of metric and tropical methods to the $\lambda$-calculus.
